Review Paper Draft

### 2020-08-13 lmm-genetic-review.tex
2020-08-13 meeting notes/edits

\begin{abstract}
Many genetic studies that aim to identify genetic variants associated with complex phenotypes are subject to unobserved confounding factors, such as environmental heterogeneity. This poses a challenge to detecting associations of interest and is known to induce spurious associations when left unaccounted for. Penalized linear mixed models (LMMs) are an attractive method to correct for unobserved confounding. These methods correct for varying levels of relatedness and population structure by modeling it as a random effect with a covariance structure estimated from observed genetic data. Despite an extensive literature on penalized regression and LMMs separately, the two are rarely discussed together. The aim of this review is to do so while offering a rigorous study of the statistical properties of penalized LMMs. Specifically, the ability of penalized LMMs to accurately estimate fixed, sparse genetic effects in the presence of environmental heterogeneity, for which population structure may serve as a proxy, has not been well studied. Additionally, the non-genetic mechanism by which population stratification may confound the phenotype-genotype relationship is rarely discussed or treated explicitly in the statistical genetics literature. Considering this, we evaluate the performance of several penalized regression methods in terms of MSE \anna{Do we want to revisit any of the precision-recall curves?} \pb {yes, let's talk about this today -- it is highly likely a reviewer will bring it up} in the presence of a number of confounding structures, comparing penalized LMMs to other approaches.
% in the presence of confounding due to varying levels of environmental heterogeneity and relatedness.
\end{abstract}

>>>>

\begin{abstract}
Many genetic studies that aim to identify genetic variants associated with complex phenotypes are subject to unobserved confounding factors, such as environmental heterogeneity. This poses a challenge to detecting associations of interest and is known to induce spurious associations when left unaccounted for. Penalized linear mixed models (LMMs) are an attractive method to correct for unobserved confounding. These methods correct for varying levels of relatedness and population structure by modeling it as a random effect with a covariance structure estimated from observed genetic data. Despite an extensive literature on penalized regression and LMMs separately, the two are rarely discussed together. The aim of this review is to do so while offering a rigorous study of the statistical properties of penalized LMMs. Specifically, the ability of penalized LMMs to accurately estimate fixed, sparse genetic effects in the presence of environmental heterogeneity, for which population structure may serve as a proxy, has not been well studied. Additionally, the non-genetic mechanism by which population stratification may confound the phenotype-genotype relationship is rarely discussed or treated explicitly in the statistical genetics literature. Considering this, we evaluate the performance of several penalized regression methods in terms of MSE \anna{Do we want to revisit any of the precision-recall curves?} \pb {yes, let's talk about this today -- it is highly likely a reviewer will bring it up} \anna{yes - work in the proportion of the first 50 vars that are correct, not necessarily a prc curve. Consider partial AUC... area under curve up to something like 50\% false discovery rate - focus on more relevant part in the path - google 'partial auc'} in the presence of a number of confounding structures, comparing penalized LMMs to other approaches.
% in the presence of confounding due to varying levels of environmental heterogeneity and relatedness.
\end{abstract}

### 2020-08-13 main.tex
2020-08-13 meeting notes/edits

The objective of many genetic studies is to accurately identify variants associated with complex traits or disease, and to estimate the effects of those variants. Such information is valuable for assessing an individual's risk of developing a particular condition, elucidating the underlying genetic architecture of a trait or disease, and identifying potential drug targets. However, this effort may be hindered by several factors such as non-independent errors \pb{is ``error'' the best term here?}, high dimensions \pb{is ``dimension'' the best term here?}, and linkage disequilibrium (LD).

\pb{This paragraph seems like too much of a digression \ldots let's get to the point sooner} Genetic variants are often correlated with each other such that univariate testing may be prone to false positives and biased estimates. When a genetic variant of interest is highly correlated with others, the effects of these will be attributed to the variant in question. \anna{I feel like I'm saying 'variant' way too much... SNP? gene?} One method for analyzing data in the presence of correlated loci is LD pruning. LD pruning can be an important preprocessing step in a GWAS analysis, for which there exist a variety of algorithms and methods. The overarching approach of these methods is to sequentially thin SNPs, so that only SNPs correlated with each other below some desired threshold are used in modeling and testing procedures. This may result in causal SNPs being removed from analysis, so that only SNPs in LD with the causal loci, as opposed to the causal loci themselves, can be identified and their effects estimated.

@@

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods \pb{I really like this sentence \ldots can we work it into the abstract?}. The remainder of this paper is organized as follows. In section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations \citep{Sillanpaeae2011, sul2018population}, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-Lasso) and lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of these methods as well as the consequences \pb{do we review \emph{consequences} in section 3?} of the ways in which they attempt to correct for population structure and environmental heterogeneity. In section \ref{sec:results}, we illustrate the concepts described in section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. These simulation results are used to illustrate the importance of distinguishing between population structure and heterogeneous environmental exposures in penalized regression methods \pb{are we still illustrating this?}. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

@@

This mechanism of environmental confounding is summarized in the directed acyclic graph displayed in figure \ref{fig:ps_env} \pb{typically, this is capitalized so it matches how Figure 1 appears in the caption}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors directly influence population-specific allele frequencies, or vice versa, only that the two may be related. In the context of our air quality example, this means that we assume air quality is not directly altering individuals' allele frequencies, nor are allele frequencies causing individuals to move to regions of better or worse air quality. Genetic relatedness and environmental effects may become associated over time for any number of complex reasons beyond the scope of this paper. However, as we will discuss, many methods of correcting for population structure rely on this relationship and do not formally differentiate between effects of population structure and those of environment. Such methods assume that genetic relatedness serves as an accurate proxy for environmental exposures, which may or may not be the case depending on the exposure and population in question. \pb{is this an important point?}

@@

An additional consideration in genetic studies that we will briefly mention is that of genetic background. \anna{make the following sentence better} Genetic background describes when SNPs that have a causal relationship with the phenotype of interest are not directly observed, but are correlated with SNPs that are observed. As \citep{vilhjalmsson2012nature} \pb{Use citet, not citep here} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. Although there are a number of methods in which to account for genetic background \anna{what are they?}, the methods upon which we will focus address this by conducting a genomewide multivariate analysis.

@@

Statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. In order to reduce confounding these methods attempt to block the causal pathway between population structure and observed SNP data. This is illustrated by the blocked arrow in figure \ref{fig:ps_env_block}. These methods, in general, do not specifically address the environmental component of confounding due to population structure. Nevertheless, they may be effective if they are able to indirectly block the causal pathway between the environment and observed SNPs. Whether or not they are effective depends upon whether population structure can be accurately inferred, and if this is a good proxy for environmental effects.

@@

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1]{figures/eta_beta_hat.png}
     \caption{$\bbeta$ and $\eta$ estimation accuracy of LMM-Lasso-Rakitsch and LMM-Lasso-ggmix for varying $\eta$ levels and sample sizes with independent subpopulation data, coarse subpopulation structure, $\xi = 0.8$, and dichotomous-discordant environmental effect structure.}
    \label{fig:eta_beta_mse}
\end{figure}

@@

@@

>>>>

\anna{get to the point sooner}

The objective of many genetic studies is to accurately identify variants associated with complex traits or disease, and to estimate the effects of those variants. Such information is valuable for assessing an individual's risk of developing a particular condition, elucidating the underlying genetic architecture of a trait or disease, and identifying potential drug targets. However, this effort may be hindered by several factors such as non-independent samples, high dimensions \pb{is ``dimension'' the best term here?} \anna{large numbers of variants >> n}, and linkage disequilibrium (LD).

\pb{This paragraph seems like too much of a digression \ldots let's get to the point sooner} \anna{this could be more like a sentence} Genetic variants are often correlated with each other such that univariate testing may be prone to false positives and biased estimates. When a genetic variant of interest is highly correlated with others, the effects of these will be attributed to the variant in question. \anna{I feel like I'm saying 'variant' way too much... SNP? gene?} One method for analyzing data in the presence of correlated loci is LD pruning. LD pruning can be an important preprocessing step in a GWAS analysis, for which there exist a variety of algorithms and methods. The overarching approach of these methods is to sequentially thin SNPs, so that only SNPs correlated with each other below some desired threshold are used in modeling and testing procedures. This may result in causal SNPs being removed from analysis, so that only SNPs in LD with the causal loci, as opposed to the causal loci themselves, can be identified and their effects estimated.

@@

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods \pb{I really like this sentence \ldots can we work it into the abstract?}. The remainder of this paper is organized as follows. In section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations \citep{Sillanpaeae2011, sul2018population}, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-Lasso) and lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of these methods as well as the consequences \pb{do we review \emph{consequences} in section 3?} of the ways in which they attempt to correct for population structure and environmental heterogeneity. In section \ref{sec:results}, we illustrate the concepts described in section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

@@

This mechanism of environmental confounding is summarized in the directed acyclic graph displayed in figure \ref{fig:ps_env} \pb{typically, this is capitalized so it matches how Figure 1 appears in the caption}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors directly influence population-specific allele frequencies, or vice versa, only that the two may be related. In the context of our air quality example, this means that we assume air quality is not directly altering individuals' allele frequencies, nor are allele frequencies causing individuals to move to regions of better or worse air quality. Genetic relatedness and environmental effects may become associated over time for any number of complex reasons beyond the scope of this paper. However, as we will discuss, many methods of correcting for population structure rely on this relationship and do not formally differentiate between effects of population structure and those of environment. Such methods assume that genetic relatedness serves as an accurate proxy for environmental exposures, which may or may not be the case depending on the exposure and population in question. \pb{is this an important point?} \anna{hmm... maybe? robust to this assumption: if it's not a good proxy it's not going to cause confounding}

@@

An additional consideration in genetic studies that we will briefly mention is that of genetic background. \anna{make the following sentence better: correlated snps are the issue - talk about ld pruning} Genetic background describes when SNPs that have a causal relationship with the phenotype of interest are not directly observed, but are correlated with SNPs that are observed. As \citet{vilhjalmsson2012nature} \pb{Use citet, not citep here} \anna{double check all citations for whether they should be (1) citet, citep or (2) better at the end of the sentence} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. Although there are a number of methods in which to account for genetic background \anna{what are they? \textbf{LD pruning} or block-wise analysis, gene-level test, \textbf{haplotype block}...group correlated variants}, the methods upon which we will focus address this by conducting a genomewide multivariate analysis.

@@

\anna{switch first and second paragraphs... the current first paragraph doesn't describe GC}

Statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. In order to reduce confounding these methods attempt to block the causal pathway between population structure and observed SNP data. This is illustrated by the blocked arrow in figure \ref{fig:ps_env_block}. These methods do not directly address the environmental component of confounding due to population structure. Nevertheless, they may be effective if they are able to indirectly block the causal pathway between the environment and observed SNPs. Whether or not they are effective depends upon whether population structure can be accurately inferred, and if this is a good proxy for environmental effects. \anna{maybe say something about how 'if it's a good proxy - great! they'll work', but 'if it's not a good proxy, also fine! because then it's not a confounder! ...SE's potentially a problem like in GLS... but this is different with lasso. (maybe not worth getting into here)}

\anna{make a better slashed arrow}

@@

\anna{in admixed data less of the environmental effect is in the column space of X so it's less of a confounder}

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1]{figures/eta_beta_hat.png}
     \caption{$\bbeta$ and $\eta$ estimation accuracy of LMM-Lasso-Rakitsch and LMM-Lasso-ggmix for varying $\eta$ levels and sample sizes with independent subpopulation data, coarse subpopulation structure, $\xi = 0.8$, and dichotomous-discordant environmental effect structure.}
    \label{fig:eta_beta_mse}
\end{figure}
\anna{add time to this figure? third row...}

@@

\anna{make new time box plot - two boxes instead of ratio - also run for n = 1000, 1200, maybe 1400 and show MSE along with time... compare convergence results to ncvfit}

@@

\anna{the importance of distinguishing between population structure and heterogeneous environmental exposures in penalized regression methods - do we want to get into the importance of not conflating environmental effects with heritability?}
